Make interactive loan word maps

### 2025-05-14 analysis/paper/loan word maps.qmd
network analysis

>>>>

## Networks
Next, we will create networks from this data, in order to be able to analyse and visualise connections.The dataset of languages is the nodelist, the edges are made on the basis of shared forms. First, the edge list is made. Fir this, we need more detailed classifications than for the maps above, so we need to read the data in again without simplifying the classifications.

```{r prepare data for network}
# read classified loan word data

# read the csv files into a list of data frames
net_data_list <- purrr::map(objs, readr::read_csv)

# Assign names
names(net_data_list) <- c("gold", "fonio", "millet", "salt", "iron")

# Function to prepare each dataset for making it into an edge list
prepare_data <- function(data) {
  data |> 
    # Step 1: Split the 'style' column into hierarchical levels
    dplyr::mutate(
      cl1 = stringr::str_extract(style, "^[A-Z]"),                    # First level
      cl2 = stringr::str_extract(style, "^[A-Z]\\.\\d+"),             # Second level
      cl3 = stringr::str_extract(style, "^[A-Z]\\.\\d+\\.\\w+")       # Third level
    ) |> 
    # Step 2: Replace "X" with NA in cl1, cl2, and cl3
    dplyr::mutate(
      cl1 = dplyr::if_else(cl1 == "X", NA_character_, cl1),
      cl2 = dplyr::if_else(cl2 == "X", NA_character_, cl2),
      cl3 = dplyr::if_else(cl3 == "X", NA_character_, cl3)
    )
}

# Apply the function to each dataset in the list
net_data_list <- lapply(net_data_list, prepare_data) 
```

```{r edge list function}
# Function to generate edge list
generate_edgelist <- function(data) {
  # Step 1: Create pairwise comparisons of glott_code values
  edgelist <- expand.grid(glott_code_1 = data$glott_code, 
                          glott_code_2 = data$glott_code,
                          stringsAsFactors = FALSE) |>
    # Remove self-loops
    dplyr::filter(glott_code_1 != glott_code_2) |>
    # Add corresponding cl1, cl2, cl3 for both nodes
    dplyr::left_join(data, by = c("glott_code_1" = "glott_code")) |>
    dplyr::rename(cl1_1 = cl1, cl2_1 = cl2, cl3_1 = cl3) |>
    dplyr::left_join(data, by = c("glott_code_2" = "glott_code")) |>
    dplyr::rename(cl1_2 = cl1, cl2_2 = cl2, cl3_2 = cl3) |>
    # Calculate weights based on equivalence
    dplyr::mutate(
      weight = dplyr::case_when(
        cl1_1 == cl1_2 & cl2_1 == cl2_2 & cl3_1 == cl3_2 ~ 3,  # Match in cl1, cl2, and cl3
        cl1_1 == cl1_2 & cl2_1 == cl2_2 ~ 2,                  # Match in cl1 and cl2
        cl1_1 == cl1_2 ~ 1,                                   # Match in cl1 only
        TRUE ~ 0                                              # No match
      )
    ) |>
    # Filter out pairs with weight = 0 (no equivalence)
    dplyr::filter(weight > 0) |>
    dplyr::select(glott_code_1, glott_code_2, weight)

  # Step 2: Reduce the edgelist to retain only one edge per pair with the highest weight
  reduced_edgelist <- edgelist |>
    # Ensure undirected edges by sorting the glott_code pair
    dplyr::mutate(
      node_1 = pmin(glott_code_1, glott_code_2),
      node_2 = pmax(glott_code_1, glott_code_2)
    ) |>
    # Group by the unique undirected pair
    dplyr::group_by(node_1, node_2) |>
    # Retain only the edge with the maximum weight
    dplyr::summarise(weight = max(weight), .groups = "drop") |>
    # Rename columns back to glott_code_1 and glott_code_2 for clarity
    dplyr::rename(glott_code_1 = node_1, glott_code_2 = node_2)

  return(reduced_edgelist)
}
```

```{r create edge lists}
edgelists <- lapply(net_data_list, generate_edgelist)
```

now, we read this into tidygraph network objects. First we take the language list and convert it into a usable edge list, and then make both into a network.
```{r}
# Convert langloc into a tibble with latitude and longitude fields
node_list <- langloc |>
  dplyr::mutate(
    latitude = sf::st_coordinates(geometry)[, 2],
    longitude = sf::st_coordinates(geometry)[, 1]
  ) |>
  tibble::as_tibble() |> 
  dplyr::select(glott_code, LanName, latitude, longitude) |>

#add some missing values
 tibble::add_row(glott_code = "dogo1299", 
                             LanName = "Dogon", 
                             latitude = 14.43, 
                             longitude = -3.49 )

# Create a list of tidygraph objects using the filtered node list
graphs <- lapply(edgelists, function(edges) {
  # Filter the node list to include only nodes present in the edge list
  filtered_nodes <- node_list |>
    dplyr::filter(glott_code %in% c(edges$glott_code_1, edges$glott_code_2))

  # Create the tidygraph object
  tidygraph::tbl_graph(
    nodes = filtered_nodes,
    edges = edges,
    directed = FALSE
  )
})

```

```{r graph plotting functions}
# Function to plot a tidygraph object with geographic layout
plot_geo_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title") {
  # Ensure that the graph has the required attributes
  if (!latitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Latitude column", latitude, "not found in node attributes."))
  }
  if (!longitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Longitude column", longitude, "not found in node attributes."))
  }
  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
    stop(paste("Weight column", weight, "not found in edge attributes."))
  }
  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Label column", label, "not found in node attributes."))
  }

  # Plot the tidygraph object
  ggraph::ggraph(graph, layout = "manual", x = !!rlang::sym(longitude), y = !!rlang::sym(latitude)) +
    ggraph::geom_edge_link(
      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
      color = "gray"
    ) +
    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
    ggraph::geom_node_point(size = 5, color = "blue") +
    ggraph::geom_node_text(
      ggplot2::aes(label = !!rlang::sym(label)), 
      repel = TRUE, 
      size = 3, 
      color = "black"
    ) +
    ggplot2::labs(title = title) +
    ggraph::theme_graph() +
    ggplot2::theme(
      legend.position = "none"
    )
}

# Function to plot a tidygraph object with force-directed layout
plot_fr_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title") {
  # Ensure that the graph has the required attributes
  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
    stop(paste("Weight column", weight, "not found in edge attributes."))
  }
  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Label column", label, "not found in node attributes."))
  }

  # Plot the tidygraph object
  ggraph::ggraph(graph, layout = "fr") +
    ggraph::geom_edge_link(
      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
      color = "gray"
    ) +
    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
    ggraph::geom_node_point(size = 5, color = "blue") +
    ggraph::geom_node_text(
      ggplot2::aes(label = !!rlang::sym(label)), 
      repel = TRUE, 
      size = 3, 
      color = "black"
    ) +
    ggplot2::labs(title = title) +
    ggraph::theme_graph() +
    ggplot2::theme(
      legend.position = "none"
    )
}

# Function to plot a tidygraph object with geographic layout and louvain communities
plot_comm_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title", subtitle = NULL, community = "louvain") {
  # Ensure that the graph has the required attributes
  if (!latitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Latitude column", latitude, "not found in node attributes."))
  }
  if (!longitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Longitude column", longitude, "not found in node attributes."))
  }
  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
    stop(paste("Weight column", weight, "not found in edge attributes."))
  }
  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Label column", label, "not found in node attributes."))
  }
   if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
    stop(paste("Label column", community, "not found in node attributes."))
  }

  # Plot the tidygraph object
  ggraph::ggraph(graph, layout = "manual", x = !!rlang::sym(longitude), y = !!rlang::sym(latitude)) +
    ggraph::geom_edge_link(
      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
      color = "gray"
    ) +
    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
    ggraph::geom_node_point(aes(colour = !!rlang::sym(community)),size = 5) +
    ggraph::geom_node_text(
      ggplot2::aes(label = !!rlang::sym(label)), 
      repel = TRUE, 
      size = 3, 
      color = "black"
    ) +
    ggplot2::labs(title = title, subtitle = subtitle) +
    ggraph::theme_graph() +
    ggplot2::theme(
      legend.position = "none"
    )
}

```

```{r}
plot_geo_graph(graphs$gold, title = "gold")
```

Improve the gold graph:
remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not.
Retain only stronger ties.

```{r}
library(tidygraph)
graphs$gold |> 
  mutate(louvain = group_louvain(weights = weight)) |> 
  mutate(louvain = as.character(louvain)) |> 
  activate(edges) |> 
  filter(weight > 1) |> 
  activate(nodes) |>
  filter(!node_is_isolated()) |>

  plot_comm_graph(community = "louvain", title = "gold", subtitle = "Louvain network communities")

```

```{r}
graphs$salt |> 
  mutate(louvain = group_louvain(weights = weight)) |> 
  mutate(louvain = as.character(louvain)) |> 
  activate(edges) |> 
  filter(weight > 1) |> 
  activate(nodes) |>
  filter(!node_is_isolated()) |>

  plot_comm_graph(community = "louvain", title = "salt", subtitle = "Louvain network communities")
```

Create a graph each with betweenness centrality
```{r betweenness graphs}
library(tidygraph)
library(ggraph)
library(dplyr)

# Iterate over each "type" in the edge list and create a graph
edge_list_summary |>
  dplyr::group_split(type) |>
  purrr::walk(\(edges_for_type) {
    # Get the type name
    type_name <- unique(edges_for_type$type)

    # Create a filtered tidygraph object for this type
    graph_for_type <- tidygraph::tbl_graph(
      nodes = node_list,
      edges = edges_for_type,
      directed = FALSE
    ) |>
      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes

    # Compute betweenness centrality for node sizes
    graph_for_type <- graph_for_type |>
      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness())

    # Plot the graph
    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
      ggraph::geom_node_point(aes(size = betweenness), colour = "blue") +
      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
      ggplot2::theme_void() +
      ggplot2::labs(
        title = paste("Graph for Type:", type_name),
        size = "Betweenness",
        width = "Edge Weight"
      ) +
      ggplot2::scale_size_continuous(range = c(2, 10)) +
      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
      ggraph::theme_graph(base_size = 12)

    print(plot)
  })
```
Create a graph each with eigenvector centrality
```{r betweenness graphs}
library(tidygraph)
library(ggraph)
library(dplyr)

# Iterate over each "type" in the edge list and create a graph
edge_list_summary |>
  dplyr::group_split(type) |>
  purrr::walk(\(edges_for_type) {
    # Get the type name
    type_name <- unique(edges_for_type$type)

    # Create a filtered tidygraph object for this type
    graph_for_type <- tidygraph::tbl_graph(
      nodes = node_list,
      edges = edges_for_type,
      directed = FALSE
    ) |>
      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes

    # Compute betweenness centrality for node sizes
    graph_for_type <- graph_for_type |>
      tidygraph::mutate(eigenvector = tidygraph::centrality_eigen())

    # Plot the graph
    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
      ggraph::geom_node_point(aes(size = eigenvector), colour = "purple") +
      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
      ggplot2::theme_void() +
      ggplot2::labs(
        title = paste("Graph for Type:", type_name),
        size = "Eigenvector",
        width = "Edge Weight"
      ) +
      ggplot2::scale_size_continuous(range = c(2, 10)) +
      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
      ggraph::theme_graph(base_size = 12)

    print(plot)
  })
```

Save a graph for each of the trade items
```{r}
# Create a list of tidygraph objects, one for each "type"
graph_list <- edge_list_summary |>
  dplyr::group_split(type) |>
  purrr::map(\(edges_for_type) {
    # Get the type name
    type_name <- unique(edges_for_type$type)

    # Create a filtered tidygraph object for this type
    graph_for_type <- tidygraph::tbl_graph(
      nodes = node_list,
      edges = edges_for_type,
      directed = FALSE
    ) |>
      tidygraph::filter(!tidygraph::node_is_isolated()) |> # Remove isolated nodes
      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness()) # Add betweenness centrality

    # Return the graph, named with the type
    graph_for_type
  }) |>
  purrr::set_names(unique(edge_list_summary$type))

# Save the list of graphs for further processing
save(graph_list, file = "graph_list.RData")

```

```{r}

```

```{r}
#knock out WOlof from the gold network
graph |> 
  activate(edges) |> 
  filter(type == "gold") |> 
  activate(nodes) |> 
  filter(LanName != "Wolof") |>
  filter(!node_is_isolated()) |> 
  plot_graph()
```

The graphs need to be sparsified now, as they are relatively useless as they are.
```{r}
library(backbone)

disp <- disparity(graph_list$gold, alpha = 0.5)
plot_graph(disp)

```

```{r}
# Find glott_code values in edge lists that are not in the node_list
find_missing_glott_codes <- function(edgelists, node_list) {
  # Combine all unique glott_code values from the edge lists
  all_glott_codes <- lapply(edgelists, function(edges) {
    unique(c(edges$glott_code_1, edges$glott_code_2))
  }) |>
    unlist() |>  # Flatten into a single vector
    unique()     # Remove duplicates

  # Find glott_codes not present in the node_list
  missing_glott_codes <- setdiff(all_glott_codes, node_list$glott_code)

  return(missing_glott_codes)
}

# Apply the function
missing_glott_codes <- find_missing_glott_codes(edgelists, node_list)

# Display the missing glott_code values
print(missing_glott_codes)

```
